Multi-target detection with application to cryo-electron microscopy

### 2019-05-24 InverseProblemsV2/inverse_problems_v2.tex
inverse problems v2

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\ii}{\textbf{i}}
\newcommand{\jj}{\textbf{j}}
\newcommand{\Poisson}{\text{Poisson}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\kk}{\textbf{k}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\mb}{\mathbf}
\newcommand{\mk}{\mathfrak}
\newcommand{\mc}{\mathcal}
\newcommand{\TODO}[1]{{\color{red}{[#1]}}}
\newcommand{\revised}[1]{{\color{blue}{#1}}}
\newcommand{\R}{\mathbb{R}}
%\newcommand{\M}{\mathcal{M}}
\newcommand{\M}{m}
\renewcommand{\P}{\mathbb{P}}
\renewcommand{\L}{\mathcal{L}}

@@

We consider the multi-target detection problem of recovering a set of signals that appear 
multiple times at unknown locations in a noisy measurement.
In the low noise regime, one can estimate the signals by first detecting occurrences, then clustering and averaging them.
In the high noise regime however, neither detection nor clustering can be performed reliably, so that strategies along these lines are destined to fail.  
Notwithstanding, using autocorrelation analysis, we show that the impossibility to detect and cluster signal occurrences in the presence of high noise does not necessarily preclude signal estimation.
Specifically, to estimate the signals, we derive simple relations between the autocorrelations of the observation and those of the signals. These autocorrelations can be estimated accurately at any noise level given a sufficiently long measurement. 
%To recover the signals from the observed autocorrelations, we (heuristically) solve a set of polynomial equations through nonlinear least-squares. 
To recover the signals from the observed autocorrelations, we  solve a set of polynomial equations through nonlinear least-squares. 
We provide analysis regarding well-posedness of the task, and demonstrate numerically the effectiveness of the method in a variety of settings.

@@

We consider the \emph{multi-target detection} problem of recovering a set of $K$ signals that appear 
multiple times at unknown locations in a noisy measurement.
Let ${x_1,\ldots,x_K\in\RL}$ be the sought signals and let $y\in\RN$ be the observed data, where we assume $N$ is  much larger than $L$. 
Let  $s[i]$ count the number of signal occurrences whose first entry is positioned at $y[i]$. Each of those $s[i]$ signals is chosen among $x_1, \ldots, x_K$ according to some (possibly unknown) distribution over $\{1,\ldots,K\}$. 
If signal occurrences overlap, they interfere additively. %We think of those events as being rare. 
With additive white Gaussian noise, the measurement model can be written as 
\begin{equation} 
	y  =  \sum_{k=1}^K s_k \ast x_k + \varepsilon, \qquad  \varepsilon   \sim \mathcal{N}(0,\sigma^2 I_N),
	\label{eq:model}
\end{equation}
where $\ast$ denotes linear convolution, and $s_k[i]$ indicates the number of occurrences of $x_k$ starting at $y[i]$, so that $s =  s_1+\cdots+s_K$. Explicitly, with zero-based indexing, % and zero-padding out of bounds,
\begin{align*}
	y[i] & = \sum_{k=1}^{K} \sum_{j = 0}^{L-1} s_k[i-j] x_k[j] + \varepsilon[i].
\end{align*}
%We discuss generative models for $s_1, \ldots, s_K$ below.
%
The goal is  to estimate $x_1,\ldots,x_K$ from $y$. % while $s$ is the \emph{nuisance variable} of the problem.
%We name this problem \emph{multi-target detection}.
In parts of the paper, we focus on the case $K = 1$, called the \emph{homogeneous} case; the case $K \geq 2$ is called \emph{heterogeneous}.
This idealized setup appears in several scientific applications, including structural biology~\cite{bendory2018toward} (as we detail below), spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive}, and system identification~\cite{ljung1998system}.

@@

\begin{figure}[t]
	\centering
	%\includegraphics[width=.7\linewidth]{KLXP/XP1}
	\includegraphics[width=.7\linewidth]{XP1}
	\caption{In the $N \to \infty$ regime (access to exact autocorrelations, excluding biased entries) and with known uniform densities, it seems that $K$ up to $\sqrt{L}$ (red curve) i.i.d.\ Gaussian signals of length $L$ can be recovered from the known moments. CPU time is in seconds. Strictly above red dots, recovery is impossible because the number of unknowns exceeds the number of computed autocorrelations; see Section~\ref{sec:heterogeneity}. Similarly to~\cite[Fig.~4.1]{boumal2017heterogeneous}, this experiment suggests a possible statistical-computational gap.}
	\label{fig:KLXP}
\end{figure}

>>>>

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\ii}{\textbf{i}}
\newcommand{\jj}{\textbf{j}}
\newcommand{\Poisson}{\text{Poisson}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\kk}{\textbf{k}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\mb}{\mathbf}
\newcommand{\mk}{\mathfrak}
\newcommand{\mc}{\mathcal}
\newcommand{\TODO}[1]{{\color{red}{[#1]}}}
\newcommand{\revised}[1]{{\color{magenta}{#1}}}
\newcommand{\R}{\mathbb{R}}
%\newcommand{\M}{\mathcal{M}}
\newcommand{\M}{m}
\renewcommand{\P}{\mathbb{P}}
\renewcommand{\L}{\mathcal{L}}

@@

We consider the multi-target detection problem of recovering a set of signals that appear 
multiple times at unknown locations in a noisy measurement.
In the low noise regime, one can estimate the signals by first detecting occurrences, then clustering and averaging them.
In the high noise regime, however, neither detection nor clustering can be performed reliably, so that strategies along these lines are destined to fail.  
Notwithstanding, using autocorrelation analysis, we show that the impossibility to detect and cluster signal occurrences in the presence of high noise does not necessarily preclude signal estimation.
Specifically, to estimate the signals, we derive simple relations between the autocorrelations of the observation and those of the signals. These autocorrelations can be estimated accurately at any noise level given a sufficiently long measurement. 
%To recover the signals from the observed autocorrelations, we (heuristically) solve a set of polynomial equations through nonlinear least-squares. 
To recover the signals from the observed autocorrelations, we  solve a set of polynomial equations through nonlinear least-squares. 
We provide analysis regarding well-posedness of the task, and demonstrate numerically the effectiveness of the method in a variety of settings.

@@

We consider the \emph{multi-target detection} problem of recovering a set of $K$ signals that appear 
multiple times at unknown locations in a noisy measurement.
Let ${x_1,\ldots,x_K\in\RL}$ be the sought signals and let $y\in\RN$ be the observed data, where we assume $N$ is  much larger than $L$. 
Let  $s[i]$ count the number of signal occurrences whose first entry is positioned at $y[i]$. Each of those $s[i]$ signals is chosen among $x_1, \ldots, x_K$ according to some (possibly unknown) distribution over $\{1,\ldots,K\}$. 
If signal occurrences overlap, they interfere additively. %We think of those events as being rare. 
With additive white Gaussian noise, the measurement model can be written as 
\begin{align} 
	y  &=  \sum_{k=1}^K s_k \ast x_k + \varepsilon, &\varepsilon   \sim \mathcal{N}(0,\sigma^2 I_N),
	\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution, and $s_k[i]$ indicates the number of occurrences of $x_k$ starting at $y[i]$, so that $s =  s_1+\cdots+s_K$. Explicitly, with zero-based indexing, % and zero-padding out of bounds,
\begin{align*}
	y[i] & = \sum_{k=1}^{K} \sum_{j = 0}^{L-1} s_k[i-j] x_k[j] + \varepsilon[i].
\end{align*}
%We discuss generative models for $s_1, \ldots, s_K$ below.
%
The goal is  to estimate $x_1,\ldots,x_K$ from $y$. % while $s$ is the \emph{nuisance variable} of the problem.
%We name this problem \emph{multi-target detection}.
In parts of the paper, we focus on the case $K = 1$, called the \emph{homogeneous} case; the case $K \geq 2$ is called \emph{heterogeneous}.
This idealized setup appears in several scientific applications, including structural biology~\cite{bendory2018toward} (as we detail below), spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive}, and system identification~\cite{ljung1998system}.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.7\linewidth]{XP1}
	\caption{In the $N \to \infty$ regime (access to exact autocorrelations, excluding biased entries) and with known uniform densities, it seems that $K$ up to $\sqrt{L}$ (red curve) i.i.d.\ Gaussian signals of length $L$ can be recovered from the known moments. CPU time is in seconds. Strictly above red dots, recovery is impossible because the number of unknowns exceeds the number of computed autocorrelations; see Section~\ref{sec:heterogeneity}. Similarly to~\cite[Fig.~4.1]{boumal2017heterogeneous}, this experiment suggests a possible statistical-computational gap.}
	\label{fig:KLXP}
\end{figure}
